---
title: "Generalised Linear Models, 2:  Transformed linear and multiple linear regression"
author: "Manazir"
date: "Copyright Loughborough University, `r format(Sys.time(), '%Y')`  \n Last update: `r format(Sys.time(), '%d %B')`"
output:
  html_document:
    self_contained: true
    highlight: textmate  # specifies the syntax highlighting style
    toc: true # should a table of contents (TOC) be shown in the document?
    toc_depth: 2 # the number of levels (e.g. section, subsection, subsubsection) shown in the TOC
    number_sections: false # should the sections be numbered?
---
# Preamble
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "",
  results = "hold",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 4,
  fig.height = 2.6,
  fig.align = "center"
)
```
## Read in packages
```{r, message = FALSE,comments=FALSE, warning = FALSE}
library("tidyverse")
library("magrittr")
library("here")
library("janitor")
library("readxl")
library("lindia")
library("rio")
```

```{r}
trees2 <- trees %>%
  mutate(Diameter = Girth) %>%
  select(Height,Diameter,Volume)
```

```{r, message = FALSE,comments=FALSE}
autot<-clean_names(read_csv(here("data", "Automobile_data.csv"),
                            na = c("?","","NA","N/A","Na"),))  #this ensures that missing data recorded in a variety of ways is identified as missing data by R
``` 

```{r}
linmod4<-lm(highway_mpg~width+curb_weight+fuel_type+body_style, data=autot)
summary(linmod4)
```
```{r}
linmod4$coefficients
```

How many fewer mpg does a car get on average for each additional pound of curb weight? For each additional inch of width?

```{r}
# for each additional pound of weight
-0.01148116
# for each additional inch of width

-0.06352332 
```

All other variables being equal, how many more mpg did wagons get on average than hatchbacks?

```{r}

# coef wagon - hatchback
1.51798186 -    0.81921943

```

What is the benefit in mpg to diesel over gas (petrol)?

```{r}
# Petrol gets 8.7 less mpg than base diesel
```

In a separate notebook, build a model of horsepower as a function of engine type, engine size and stroke. Invent some questions and answer them using your model.

```{r}
autot <- autot %>%
  mutate_at(vars(engine_type),
             list(factor))
```
## Ex1 Filtering
```{r}
hatchbacks <- autot %>%
  filter(body_style == "hatchback")
```


```{r}
mpgbody <- autot %>%
  group_by(body_style) %>%
  summarise(meanhw_mpg =mean(highway_mpg),
            sdhw_mpg = sd(highway_mpg),
            meancity_mpg = mean(city_mpg),
            sdcity_mpg = sd(city_mpg))
```


```{r}
autot_na <- autot%>%
  select_if(~ !any(is.na(autot)))
select(complete.cases(autot))
```
```{r}
library(rio)
```



```{r}
PL_20_21<- import("http://www.football-data.co.uk/mmz4281/2021/E0.csv")
```



```{r}
pl_20_21 <- as_tibble(PL_20_21) %>%
  select(Div:AR)
  pl_20_21$Date <- as.Date(pl_20_21$Date, format = "%d/%m/%Y")

  
```
```{r}
pl_20_21 <- pl_20_21 %>%
  mutate(season = "2020/21")
```

```{r}
HomeT <- pl_20_21 %>%
  select(Div, HomeTeam, AwayTeam, FTHG, FTAG, Date, season) %>%
  mutate(Home = 1)
```


```{r}
HomeT <- HomeT %>%
  rename(Team = HomeTeam, Opponent = AwayTeam, TeamGoals = FTHG, OpponentGoals = FTAG)
```

``` {r}
AwayT <- pl_20_21 %>%
  select(Div, HomeTeam, AwayTeam, FTHG, FTAG, Date, season) %>%
  mutate(Home = 0)
```


```{r}
AwayT <- AwayT %>%
  rename(Team = AwayTeam, Opponent = HomeTeam, TeamGoals = FTAG, OpponentGoals = FTHG)

```

```{r}
pl_poissondata <- rbind(HomeT, AwayT)
pl_poissondata[c(1,381),]

```

```{r}
poisson_pl <- glm(TeamGoals ~ Home + Team + Opponent, data = pl_poissondata, family = poisson(link ="log"))
summary(poisson_pl)
```



